Comparing ChatGPT-4 to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6942 Deaths in Sierra Leone from 2019-2022

### 2024-01-25 manuscript/manuscript.tex
Sanitized counts

Initially, 11,920 records from the HEAL-SL study \cite{njalauniversityHealthySierraLeone2023,carshon-marshChildMaternalAdult2022} were collected from dual-coded EVA, where each record was randomly coded by two different physicians that assigned CODs as International Classification of Diseases Revision 10 (ICD-10) codes \cite{worldhealthorganizationICD10InternationalStatistical2011}. Physicians were able to assign CODs for 11,820 of the 11,920 records, where 100 of these records could not be assigned a COD due to missing or inadequate information (e.g. low quality narrative, data loss). To determine if two codes were in physician agreement per record, codes were compared for similarity using Central Medical Evaluation Agreement 10 (CMEA-10) codes, which groups a range of similar ICD-10 codes together indicating codes in agreement \cite{aleksandrowiczPerformanceCriteriaVerbal2014} (see Additional File 1). When codes were not in agreement, a record enters the reconciliation phase, where the two physicians were provided reasoning and initial codes from each other to: (1) keep their initial code (2) assign the other physician's code or (3) assign a new code. If codes were not in agreement after the reconciliation phase, a record enters the adjudication phase, where a third senior physician evaluates both physicians' reasoning and codes before and after reconciliation, and assigns a final code based on their evaluation. The 11,820 physician coded records were further filtered for records where both physicians agreed on the assigned codes (records that were not reconciled or adjudicated), resulting in 6942 physician agreed records, where . Since computer models were compared to physicians in this study, there was more certainty that COD assignments agreed by both physicians were representative of physician assignment than when they disagreed \cite{barnettComparativeAccuracyDiagnosis2019,morrisFactorsAssociatedPhysician2010,hsiaoFactorsAssociatedPhysician2012}.

@@

Four computer models were used to assign COD for each of the 6942 physician agreed records: ChatGPT-3.5, ChatGPT-4, InterVA-5 , and InSilicoVA. Each model required pre-processing of the 6942 physician agreed records into input data, and standardization of output COD codes from models for performance evaluation as not all models produced comparable codes across outputs. Although each model can assign multiple CODs per record, only the first generated COD response from ChatGPT-3.5 and ChatGPT-4, and the most probable COD from InterVA-5 and InSilicoVA were used for evaluation. After conversion of all model and physician outputs to generalized Centre for Global Health Research 10 (CGHR-10) codes (see Additional File 2), 6939 records were comparable to evaluate performances of models for COD assignment relative to physicians.

@@

For ChatGPT-3.5 and ChatGPT-4, text prompts were generated for each record as input to instruct the models to assign CODs based on the open narratives. Parameter \verb|t_prob| was set to 0.3. Prompts were generated with a template involving \verb|<narrative>| from the free-text open narratives, and associated \verb|<age>|, (age) \verb|<unit>|, \verb|<sex>| from the standardized questionnaire:

@@

For InterVA-5 and InSilicoVA, the standardized questionnaire data from the HEAL-SL EVA were first converted into 2016 World Health Organization (WHO) VA questionnaire revision 1.5.1 
\cite{} Open Data Kit (ODK) format \cite{worldhealthorganizationODKVerbalAutopsy2022,nafundiODKCollectData2023}, followed by further conversion into OpenVA format \cite{liOpenVAToolkitVerbal2023} using the \verb|pyCrossVA| Python package \cite{thomasPycrossvaPrepareData}, and used as input for InterVA-5 and InSilicoVA. The \verb|Nsim| (number of iterations to run) parameter for InSilicoVA was set to 10,000, while the \verb|HIV| (level of prevalence of human immunodeficiency virus) and \verb|Malaria| (level of prevalence of Malaria) parameters were both set to \verb|'h'| (high).

The 6942 physician agreed records were all converted into instructional text prompts for ChatGPT-3.5 and ChatGPT-4. Only 6897 of the 6942 physician agreed records were converted into OpenVA formatted records for InterVA-5 and InSilicoVA, where 45 records caused conversion errors and were skipped or removed.

@@

Of the 6942 physician agreed records, ChatGPT-3.5 and ChatGPT-4 were able to assign CODs for 6932 and 6912 records respectively, while InterVA-5 and InSilicoVA both assigned CODs for 6833 records. All 6833 records with WHO VA 2016 v1.5 codes \cite{worldhealthorganizationVerbalAutopsyStandards2016} from InterVA-5 and InSilicoVA output were converted into ICD-10 codes. After all model outputs contained ICD-10 codes, they were further converted to CGHR-10 codes that generalized ICD-10 codes into 19, 10, and 7 categories for the adult (12 years or older), child (28 days to 11 years), and neonatal (under 28 days) age groups (see Additional file 2). The 6932 and 6912 records with ICD-10 codes from ChatGPT-3.5 and ChatGPT-4 output were converted into CGHR-10 codes, resulting in 6894 and 6857 records, where 38 and 55 records did not have matching CGHR-10 codes respectively. The 6833 records with ICD-10 codes from InterVA-5 and InSilicoVA output were also converted into CGHR-10 codes, resulting in and 6805 and 6729 records, where 28 and 104 records did not have matching CGHR-10 codes respectively. The 6942 physician agreed records with COD assignments in ICD-10 were also converted to CGHR-10 codes, which resulted in 6939 physician agreed records for comparison across physician and model assignments (three records in ICD-10 were not captured by the CGHR-10 ranges).

@@

The performance of four models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, were evaluated with metrics on the population and individual level by comparing their CGHR-10 COD outputs to 6939 physician agreed records. Cause Specific Mortality Fraction (CSMF) accuracy was used to evaluate models on the population level, while Partial Chance Corrected Concordance (PCCC) was used to evaluate models on the individual level \cite{murrayRobustMetricsAssessing2011}. Records that were assigned a COD by physicians, but not by a model were considered to be an incorrect COD assignment by the model. CSMF accuracy and PCCC were calculated for each model by three age groups, and further into age ranges and cause of death for each age group.

@@

The CSMF accuracy and PCCC metrics were calculated and compared for each model overall and by age group, followed by age ranges and cause of death for each age group as model performance can vary across ages and specific causes \cite{benaraEvaluationMethodsAssigning2023,murrayUsingVerbalAutopsy2014,setelValidityVerbalAutopsy2006}. Metrics were calculated overall for three age groups: adult (12 years or older), child (28 days to 11 years), and neonatal (under 28 days). For each of the adult and child age groups, metrics were calculated for five-year age ranges for records with ages at death of one-year or older and five-month age ranges for 28 days or older. For the neonatal age group, the age ranges of 0-6 days and 7-27 days were used. Metrics were also calculated by CODs defined by CGHR-10 codes, which include 19, 10, and 7 CODs for adult, child, and neonatal records respectively.

@@

\subsection{Age and COD Distributions}\label{results-distrib}

@@

\subsection{Overall Performance by Age Group}\label{results-overall}

@@

\subsection{Performance for Adult Records}\label{results-adult}

@@

\subsection{Performance for Child Records}\label{results-child}

\subsection{Performance for Neonatal Records}\label{results-neo}

>>>>

Initially, 11,920 records from the HEAL-SL study \cite{njalauniversityHealthySierraLeone2023,carshon-marshChildMaternalAdult2022} were collected from dual-coded EVA, where each record was randomly coded by two different physicians that assigned CODs as International Classification of Diseases Revision 10 (ICD-10) codes \cite{worldhealthorganizationICD10InternationalStatistical2011}. Physicians were able to assign CODs for 11,820 of the 11,920 records, where 100 of these records could not be assigned a COD due to missing or inadequate information (e.g. low quality narrative, data loss). To determine if two codes were in physician agreement per record, codes were compared for similarity using Central Medical Evaluation Agreement 10 (CMEA-10) codes, which groups a range of similar ICD-10 codes together indicating codes in agreement \cite{aleksandrowiczPerformanceCriteriaVerbal2014} (see Additional File 1). When codes were not in agreement, a record enters the reconciliation phase, where the two physicians were provided reasoning and initial codes from each other to: (1) keep their initial code (2) assign the other physician's code or (3) assign a new code. If codes were not in agreement after the reconciliation phase, a record enters the adjudication phase, where a third senior physician evaluates both physicians' reasoning and codes before and after reconciliation, and assigns a final code based on their evaluation. The 11,820 physician coded records were further filtered for records where both physicians agreed on the assigned codes (records that were not reconciled or adjudicated) resulting in 6942 physician agreed records. Since computer models were compared to physicians in this study, there was more certainty that COD assignments agreed by both physicians were representative of physician assignment than when they disagreed \cite{barnettComparativeAccuracyDiagnosis2019,morrisFactorsAssociatedPhysician2010,hsiaoFactorsAssociatedPhysician2012}. The 6942 records were converted into Centre for Global Health Research 10 (CGHR-10) codes (see Additional File 2) that generalized ICD-10 codes into 19, 10, and 7 categories for the adult (12 years or older), child (28 days to 11 years), and neonatal (under 28 days) age groups.  After conversion, a final total of 6939 physician agreed records (3826 adult, 2636 child, and 477 neonatal) was used for modelling and performance evaluation, where three records were removed as their ICD-10 codes did not have a matching CGHR-10 code.

@@

Four computer models were used to assign COD for each of the 6942 physician agreed records: ChatGPT-3.5, ChatGPT-4, InterVA-5 , and InSilicoVA. Each model required pre-processing of the 6939 records into input data, and standardization of output COD codes from models for performance evaluation as not all models produced comparable codes across outputs. Although each model can assign multiple CODs per record, only the first generated COD response from ChatGPT-3.5 and ChatGPT-4, and the most probable COD from InterVA-5 and InSilicoVA were used for evaluation. All model outputs were converted to CGHR-10 codes to evaluate performances of models for COD assignment relative to physicians.

@@

For ChatGPT-3.5 and ChatGPT-4, 6939 text prompts were generated for each physician agreed record as input to instruct the models to assign CODs based on the open narratives. Parameter \verb|t_prob| was set to 0.3. Prompts were generated with a template involving \verb|<narrative>| from the free-text open narratives, and associated \verb|<age>|, (age) \verb|<unit>|, \verb|<sex>| from the standardized questionnaire:

@@

For InterVA-5 and InSilicoVA, the standardized questionnaire data from the HEAL-SL EVA were first converted into 2016 World Health Organization (WHO) VA questionnaire revision 1.5.1 Open Data Kit (ODK) format \cite{worldhealthorganizationODKVerbalAutopsy2022,nafundiODKCollectData2023} consisting of 526 variables \cite{dipasqualeReleaseODK20162016}, followed by further conversion into OpenVA format \cite{liOpenVAToolkitVerbal2023} consisting of 353 variables \cite{byassInterVA5UserGuide2020} using the \verb|pyCrossVA| Python package \cite{thomasPycrossvaPrepareData}. The \verb|Nsim| (number of iterations to run) parameter for InSilicoVA was set to 9500 (the default of 10,000 resulted in errors and ran successfully until 9500 iterations), while the \verb|HIV| (level of prevalence of human immunodeficiency virus) and \verb|Malaria| (level of prevalence of Malaria) parameters were both set to \verb|'h'| (high). The 6939 records were all converted into OpenVA formatted records for InterVA-5 and InSilicoVA.

@@

Of the 6939 input records, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA were able to assign CODs for 6929, 6909, 6805, 6729 records respectively. All 6805 InterVA-5 and 6729 InSilicoVA records with WHO VA 2016 v1.5 output codes \cite{worldhealthorganizationVerbalAutopsyStandards2016} were converted into ICD-10 codes. After all model outputs used ICD-10 codes, they were further converted to CGHR-10 codes. The 6929 ChatGPT-3.5 and 6909 ChatGPT-4 output records with ICD-10 codes were converted into 6892 and 6856 records with CGHR-10 codes, where 37 and 53 records did not have matching CGHR-10 codes respectively. All 6805 InterVA-5 and 6729 InSilicoVA records with ICD-10 codes were converted into records with CGHR-10 codes.

@@

The performance of four models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, were evaluated with metrics on the population and individual level by comparing their CGHR-10 COD outputs to 6939 records. Cause Specific Mortality Fraction (CSMF) accuracy was used to evaluate models on the population level, while Partial Chance Corrected Concordance (PCCC) was used to evaluate models on the individual level \cite{murrayRobustMetricsAssessing2011}. Records that were assigned a COD by physicians, but not by a model were considered to be an incorrect COD assignment by the model. CSMF accuracy and PCCC were calculated for each model by three age groups, and further into age ranges and cause of death for each age group.

@@

The CSMF accuracy and PCCC metrics were calculated and compared for each model overall and by age group, followed by age ranges and cause of death for each age group as model performance can vary across ages and specific causes \cite{benaraEvaluationMethodsAssigning2023,murrayUsingVerbalAutopsy2014,setelValidityVerbalAutopsy2006}. Metrics were calculated overall for three age groups according to the CGHR-10 codes: adult (12 years or older), child (28 days to 11 years), and neonatal (under 28 days). For each of the adult and child age groups, metrics were calculated for five-year age ranges for records with ages at death of one-year or older and five-month age ranges for 28 days or older. For the neonatal age group, the age ranges of 0-6 days and 7-27 days were used. Metrics were also calculated by CODs defined by CGHR-10 codes, which include 19, 10, and 7 CODs for adult, child, and neonatal records respectively.

@@

\subsection{Overall Performance by Age Groups}\label{results-overall}

@@

\subsection{Performance for Adult Records (12 Years or Older)}\label{results-adult}

@@

\subsection{Performance for Child Records (28 Days to 11 Years)}\label{results-child}

@@

\subsection{Performance for Neonatal Records (Under 28 Days)}\label{results-neo}
